Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks

### 2021-11-17 paper/ch_potentials.tex
polishing

\begin{abstract}
We study exact, efficient and practical algorithms for route planning applications in large road networks.
Routing applications often require integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the exact application.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many applications.
A* is a classical approach to accelerate Dijkstra's algorithm.
A* can support many extended scenarios without much additional implementation complexity.
However, A*'s performance depends on the availability of a good heuristic that estimates distances.
Computing tight distance estimates is a challenge on its own.
On road networks, shortest paths can also be quickly computed using hierarchical speedup techniques.
They achieve speed and exactness but sacrifice A*'s flexibility.
Extending them to certain practical applications can be hard.
In this paper, we present Lazy RPHAST, an algorithm to efficiently compute many distance estimates towards a single target and use it as an A* heuristic.
As it uses Contraction Hierarchies (CH), a hierarchical technique, we call our A* heuristic CH-Potentials.
Our approach allows decoupling the supported applications from the hierarchical speed-up technique.
Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks and present a new pruning criterion for symmetrical bidirectional A*.
An extensive experimental study confirms the practicality of our approach for many applications.
\end{abstract}

@@

The rest of the article is organized as follows.
In Section~\ref{sec:related_work}, we discuss related works on goal directed search and extensions for realistic applications for hierarchical techniques.
After establishing preliminaries in Section~\ref{sec:preliminaries}, we introduce the incremental many-to-one problem and our algorithm to solve it efficiently in Section~\ref{sec:lazy-rphast}.
CH-Potentials, our new distance estimation function is introduced in Section~\ref{sec:main-algo}.
There, we also discuss our improvements to the handling of low-degree nodes (Section~\ref{sec:low-deg-improvment}) and for symmetrical bidirectional A* (Section~\ref{sec:bidir_astar}).
% Section~\ref{sec:extensions} presents a multitude of applications of CH-Potentials.
In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing how to apply the approach to different practical applications.
Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.

@@

A CH is a two phase technique to efficiently compute exact, shortest paths.
For details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
In this section, we give an introduction.

@@

In a CH, each node gets a unique rank representing its ``importance''.
By convention, important nodes are usually depicted at the ``top'' while unimportant nodes are positioned at the ``bottom''.
The intuition is that dead-ends are unimportant while highways or bridges are important.
An edge goes \emph{up} when its tail node has lower rank than its head node.
\emph{Down} edges are defined analogously.
An \emph{up-down path} is a path where only one node $m$ is more important than both its neighbors.
$m$ is called the \emph{mid} node.
An \emph{up path} is a path where the last node is the mid node.
Similarly, the first node is the mid node of a \emph{down path}.
% Every up and down path is an up-down path.
%
In the preprocessing phase, first, a node ordering is computed heuristically.
Then, \emph{shortcut} edges are added to the input graph $G$ to obtain $G^+$.
This is done by repeatedly contracting the least important node and adding shortcuts between its neighbors.
%See~\cite{gssv-erlrn-12} for the details.
After the preprocessing, for every pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for a proof sketch.
From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) can be constructed.
Thus, we can restrict our search to up-down paths in $G^+$.
The search is bidirectional.
The forward search starts from $s$ and only follows up-edges.
Similarly, the backward search starts at $t$ and only follows down-edges in reversed direction.
The two searches meet at the mid node.
Pseudocode for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.
%
A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
On road networks, this is the case~\cite{gssv-erlrn-12,dgpw-crprn-13}.
On graphs with low treewidth, this is also the case~\cite{dsw-cch-15,hs-gbpo-18}.

@@

Some hierarchical techniques were extended to support a three-phase setup.
The phases consist of a slow preprocessing phase, a faster \emph{customization} phase, and fast queries.
The preprocessing phase typically takes hours, the customization seconds, and the query milliseconds.
The preprocessing phase is independent of the weights and only exploits the graph topology.
Weights are introduces in the customization phase.
Paths are computed in the query phase.
This setup is useful to allow for weight changes that are common to all vehicles such as life traffic updates.
The first technique to follow this setup is CRP~\cite{dgpw-crprn-13}.

Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) are a variant of Contraction Hierarchies that uses this three-phase setup.
The key idea is to use a node importance order that is independent of the weight function.
A CCH is also a CH.
The final result of the preprocessing is also an augmented graph $G^+$.
It fulfills all necessary properties for CH query algorithms.
The CH query algorithms can therefore be applied without modifications.
However, the CCH preprocessing grants some additional useful properties.

@@

We can also skip some degree three nodes.
Denote by $u, v_1,\ldots, v_k, w$ a degree two chain as described in the previous section.
If $d(w) > 3$ or $w$ is in the queue, our algorithm proceeds as in the previous section.
Otherwise, there exist up to two degree chains $w, x_1,\ldots,x_p, y$ and $w,a_1,\ldots,a_q,b$ such that $x_1\neq y_k \neq a_1$.
Our algorithm iteratively tries to reduce all $D[x_i]$ and $D[a_i]$.
If it reaches $b$, $b$'s weight in the queue is set to $D[b]+h_t(b)$.
Analogously, if $y$ is reached, $y$'s weight is set to $D[y]+h_t(y)$.
If $y$ respectively $b$ are not reached, our algorithm does nothing.

@@

The problem with this straightforward approach is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
Thus, in the equivalent bidirectional Dijkstra search each direction would run on a different graph.
This breaks the usual bidirectional Dijkstra stopping criterion.
To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for one direction.
The forward search can skip nodes already settled by the backward search and vice versa.
However, this straightforward bidirectional A* still performs more work than a unidirectional A*~\cite{gh-cspas-05}.
Thus, on its own it is not a useful algorithm.
However, it can serve as basis for further algorithmic refinements.
The authors of~\cite{gh-cspas-05} refer to this approach as \emph{symmetric} bidirectional A*.

To obtain a bidirectional stopping criterion, the \emph{average potential} is proposed~\cite{gh-cspas-05}.
It combines a forward and a backward heuristic $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$ into a combined \emph{average heuristic} $h_{\overline{st}}$.
The idea is to use a common reduced graph, whose weights are the average weights of the individual reduced graphs.
Both searches run on the same common reduced graph.
This allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
Formally, $h_{\overline{st}}(v)$ is defined as $(\overrightarrow{h}_t(v) - \overleftarrow{h}_s(v))/2$.
The forward search uses $h_{\overline{st}}(v)$ as its heuristic.
The backward search uses $-h_{\overline{st}}(v)$.
Unfortunately, average potentials have two downsides.
First, evaluating the average potential requires evaluating both $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
Evaluating the average heuristic is therefore slower than evaluating just one of the composing heuristics.
Second, the bidirectional stopping criterion comes at the cost of worse estimates for each direction on its own.
$h_{\overline{st}}(v)$ is a worse estimate for $\dist(v,t)$ than $\overrightarrow{h}_t$.
Similarly, $-h_{\overline{st}}(v)$ is a worse estimate for $\dist(s,v)$ than $\overleftarrow{h}_s$.
The second downside can be partially mitigated through pruning with the composing heuristics.
When the forward search scans an edge $uv$ where $\overrightarrow{D}[u] + w(uv) + \overrightarrow{h}_s(v) > \mu$ holds, i.e. the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, it is not necessary to push $y$ into the queue.
The pruning rule for the backward search is analogous.

@@

To enable quick shortest path computations, we consider a two phase setup with an additional off-line preprocessing phase before the on-line query phase.
The input to the preprocessing phase is a graph $G_\ell$ with lower bound weights $w_\ell$ and a node mapping function $\phi$.
$w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
We require that $w_q(u v)$ is greater or equal to the shortest distance $\dist_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
The output of the preprocessing is auxiliary data that enables an efficient heuristic function $h_t(x)$.
$h_t(x)$ is the exact distance from $\phi(x)$ to $\phi(t)$ in $G_\ell$.
In the applications considered in this paper, $w_\ell$ is always the freeflow travel time.

@@

To combine live traffic with predicted traffic, we define a modified travel time function $w_q$ that is then used as query weights.
Denote by $w_p(e,\tau)$ the predicted travel time along edge $e$ at moment $\tau$.
Further, $w_c(e)$ is the travel time according to current live traffic.
Finally, we denote by $\tau_{\mathrm{soon}}$ the moment when we switch to predicted traffic.
In our experiments, we set $\tau_{\mathrm{soon}}$ to one hour in the future.
We need to make sure that the modified travel time function fulfills the no-waiting property.
For this reason, we cannot make a hard switch at $\tau_{\mathrm{soon}}$.
Our modified travel time function linearly approaches the predicted travel time. % with a bounded slope.
%
Formally, we set $w_q(e,\tau)$ to $w_c(e)$, if $\tau \leq \tau_{\mathrm{soon}}$.
Otherwise, we check whether $w_p(e,\tau_{\mathrm{soon}}) < w_c(e)$ is true.
If it is the case, we set $w_q(e,\tau)$ to $\max\{w_c(e)+(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
Otherwise, we set $w_q(e,\tau)$ to $\min\{w_c(e)-(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
% We then run the algorithm of Section~\ref{sec:predicted-traffic} with $w_q$ as query weight.
In our implementation, we to not modify the representation of $w_p$ but evaluate the formulas above at each travel time evaluation.
We set $w_\ell$ again to the freeflow travel time.

@@

This setup provides an alternative way of integrating live traffic compared to the approach described in Section~\ref{sec:live-predicted-traffic}.
Changes to the live traffic situation are handled by changing $w_\ell$ via a customization.
Additional features such as turn restrictions can be implemented by adjusting $w_q$.
For severe traffic events, the three-phase approach will be significantly faster than the one described in Section~\ref{sec:live-predicted-traffic}.
However, it comes at the expense of a more complex algorithm.
Further, integrating predicted traffic becomes difficult, as $w_q$ may not drop below the $w_\ell$ with integrated congestion.

@@

When looking at total running times for a single many-to-one query including selection times, Lazy RPHAST delivers very competitive performance and is as fast if not faster than all algorithms evaluated in~\cite{delling_et_al:OASIcs:2011:3266}.
For example, Delling et al. report an average RPHAST running time of 1.97\,ms for selection and query combined for $|B| = 2^{14}$ and 28.52\,ms for $|B| = 2^{20}$ on DIMACS Europe.
Lazy RPHAST (on a CH) takes 0.96\,ms and 10.76\,ms respectively, to compute the same distances.
The CCH variant is slightly slower (1.00\,ms and 14.41\,ms) but still significantly faster than RPHAST.
We conclude that that Lazy RPHAST is a very useful extension to RPHAST.
It allows efficiently handling dynamic source sets and is even faster than RPHAST in the setting where both the target and the source set change between queries.
% This is of course not a fair comparison since the approaches in~\cite{delling_et_al:OASIcs:2011:3266} are geared towards having one expensive selection phase and than many faster queries.
% When computing distances from different sources towards the same target set, RPHAST will outperform Lazy RPHAST.
% However, whenever the target set may change between queries, Lazy RPHAST offers a simple and fast alternative.

@@

Table~\ref{tab:building_blocks} depicts the performance of unidirectional A* with different heuristics and optimizations.
We compare (C)CH-Potentials to three other heuristics.
The first heuristic is, the \emph{Zero} heuristic where $h(v)$ is $0$ for all nodes $v$.
This corresponds to using Dijkstra's algorithm.
Secondly, we compare against our own implementation of ALT~\cite{gw-cppsp-05}.
We use 16 landmarks generated with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
Finally, we compare against a hypothetical \emph{Oracle-A*} heuristic.
This heuristic has instant access to a shortest distance array with respect to $w_\ell$, i.e. it is faster than the fastest heuristic possible in our model.
We fill this array before each query using a reverse Dijkstra search from the target node.
In this experiment, we do not account for the running time of Dijkstra's algorithm.
Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
(C)CH-Potentials compute the same distance estimates, but the heuristic evaluation has some overhead.
Comparing against Oracle-A* allows us to measure this overhead.
No other heuristic, which only has access to the preprocessing weights, can be faster than Oracle-A*.

@@

In this section, we investigate the performance of bidirectional A*.
We first evaluate different variants of bidirectional A* in Table~\ref{tab:bidir_pruning} and~\ref{tab:bidir_switching} then compare the best ones against unidirectional A* in Table~\ref{tab:bidir}.
Table~\ref{tab:bidir_pruning} depicts the performance of different variants of bidirectional A* with and without our improved pruning.
As observed in the previous section, enabling the low degree optimization achieves a speed-up by roughly a factor of three.
Symmetric bidirectional A* without our improved pruning has the worst performance.
Enabling the improved pruning significantly improves the performance of symmetric bidirectional A*.
For all heuristics except ALT, symmetric A* with the improved pruning has smaller search spaces than the average potential and very similar running times.
Without the low degree improvements, the improved symmetric variant is marginally faster and with the low degree improvements the average potential remains slightly faster.
This is due to the reduced impact of the heuristic evaluation overhead with the low degree optimizations.
Enabling the improved pruning for the average potential marginally reduces the search space size and slightly increases running times.

@@

In Table~\ref{tab:bidir} we investigate the effectiveness of bidirectional search compared to unidirectional search depending on the query weights.
Interestingly, only the zero heuristic and ALT consistently achieve speed-ups through bidirectional search on all query weights.
With $w_q = w_{\ell}$, unidirectional CH-Potentials are already optimal and only traverse the shortest path.
Here, bidirectional search will only introduce unnecessary overhead.
When query weights are scaled up uniformly, bidirectional search achieves some search space reduction, but it is not enough to significantly reduce running times due to the overhead of running a second search.
This changes drastically when the query weight increases are applied non-uniformly in the third scenario.
Here only weights with a speed less than 80\,kph were scaled up.
This touches only the beginning and end of most shortest paths between randomly chosen nodes.
The middle part of the shortest paths will typically use faster edges like highways.
In this case, bidirectional (C)CH-Potentials are a factor of two faster than the unidirectional variant.
The reason for this is that the search space of a unidirectional search expands greatly while exploring the end of the path to the target where the reduced weights are bad.
In contrast, the bidirectional searches meet in the middle of the shortest path where the reduced weights are close to zero and thus avoid this expansion.
This is also the reason why ALT behaves like this for all query weights.
By construction, the ALT heuristic has better reduced weights for edges which are on many shortest paths like highways.
Conversely, unimportant edges have bad reduced weights.
This makes bidirectional search so critical for ALTs performance.
In contrast, having a potential as tight as CH-Potentials makes bidirectional search in many scenarios unnecessary.
Bidirectional search for CH-Potentials only pays of when the reduced weights exhibit certain very specific patterns.

@@

Table~\ref{tab:applications} depicts the running times of (C)CH-Potentials in various applications, such as those described in Section~\ref{sec:extensions}.
We report speedups compared to extensions of Dijkstra's algorithm for each application respectively.
We start with the base case where $w_q = w_\ell$.
This is the problem variant solved by the basic CH algorithm.
CH achieves average query running times of 0.16\,ms on OSM Ger.
CH-Potentials are roughly four times slower but still achieve a huge speedup of 3795 over Dijkstra.
Such large speedups are typical for CH.
This shows that CH-Potentials gracefully converges toward a CH in the $w_q = w_\ell$ special case.
On DIMACS Europe, the average degree is somewhat higher which makes the low-degree optimizations less effective which leads to more queue operations and a slightly slower running time.

>>>>

\begin{abstract}
We study exact, efficient and practical algorithms for route planning applications in large road networks.
Routing applications often require integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the exact application.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many applications.
A* is a classical approach to accelerate Dijkstra's algorithm.
A* can support many extended scenarios without much additional implementation complexity.
However, A*'s performance depends on the availability of a good heuristic that estimates distances.
Computing tight distance estimates is a challenge on its own.
On road networks, shortest paths can also be quickly computed using hierarchical speedup techniques.
They achieve speed and exactness but sacrifice A*'s flexibility.
Extending them to certain practical applications can be hard.
In this paper, we present Lazy RPHAST, an algorithm to efficiently compute distances from many nodes towards a single target and use it as an A* heuristic.
As it uses Contraction Hierarchies (CH), a hierarchical technique, we call our A* heuristic CH-Potentials.
Our approach allows decoupling the supported applications from the hierarchical speed-up technique.
Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks and present a new pruning criterion for symmetrical bidirectional A*.
An extensive experimental study confirms the practicality of our approach for many applications.
\end{abstract}

@@

The rest of the article is organized as follows.
In Section~\ref{sec:related_work}, we discuss related works on goal directed search and extensions for realistic applications for hierarchical techniques.
After establishing preliminaries in Section~\ref{sec:preliminaries}, we introduce the incremental many-to-one problem and Lazy RPHAST, our algorithm to solve it efficiently, in Section~\ref{sec:lazy-rphast}.
CH-Potentials, our new A* heuristic is introduced in Section~\ref{sec:main-algo}.
There, we also discuss our improvements to the handling of low-degree nodes (Section~\ref{sec:low-deg-improvment}) and for symmetrical bidirectional A* (Section~\ref{sec:bidir_astar}).
% Section~\ref{sec:extensions} presents a multitude of applications of CH-Potentials.
In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing how to apply the approach to different practical applications.
Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.

@@

A CH is a two-phase technique to efficiently compute exact, shortest paths.
For details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
In this section, we give an introduction.

@@

In a CH, each node gets a unique rank representing its ``importance''.
By convention, important nodes are usually depicted at the ``top'' while unimportant nodes are positioned at the ``bottom''.
The intuition is that dead-ends are unimportant while highways or bridges are important.
An edge goes \emph{up} when its tail node has lower rank than its head node.
\emph{Down} edges are defined analogously.
An \emph{up-down path} is a path where only one node $m$ is more important than both its neighbors.
$m$ is called the \emph{mid} node.
An \emph{up path} is a path where the last node is the mid node.
Similarly, for a \emph{down path}, the first node has to be the mid node.
% Every up and down path is an up-down path.
%
In the preprocessing phase, first, a node ordering is computed heuristically.
Then, \emph{shortcut} edges are added to the input graph $G$ to obtain $G^+$.
This is done by repeatedly contracting the least important node and adding shortcuts between its higher ranked neighbors.
%See~\cite{gssv-erlrn-12} for the details.
After the preprocessing, for every pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for a proof sketch.
From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) can be constructed.
Thus, we can restrict our search to up-down paths in $G^+$.
The search is bidirectional.
The forward search starts from $s$ and only follows up-edges.
Similarly, the backward search starts at $t$ and only follows down-edges in reversed direction.
The two searches meet at the mid node.
Pseudocode for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.
%
A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
On road networks, this is the case~\cite{gssv-erlrn-12,dgpw-crprn-13}.
On graphs with low treewidth, this is also the case~\cite{dsw-cch-15,hs-gbpo-18}.

@@

Some hierarchical techniques were extended to support a three-phase setup.
The phases consist of a slow preprocessing phase, a faster \emph{customization} phase, and fast queries.
The preprocessing phase typically takes hours, the customization seconds, and the query milliseconds.
The preprocessing phase is independent of the weights and only exploits the graph topology.
Weights are introduced in the customization phase.
Paths are computed in the query phase.
This setup is useful to allow for weight changes that are common to all vehicles such as life traffic updates.
The first technique to follow this setup is CRP~\cite{dgpw-crprn-13}.

Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) are a variant of Contraction Hierarchies that uses this three-phase setup.
The key idea is to use a node importance order that is independent of the weight function.
A CCH is also a CH.
The final result of the preprocessing is also an augmented graph $G^+$.
It fulfills all necessary properties for CH query algorithms.
The CH query algorithms can therefore be applied without modifications.
However, the CCH preprocessing grants some useful additional properties.

@@

We can also skip some degree three nodes.
Denote by $u, v_1,\ldots, v_k, w$ a degree two chain as described in the previous section.
If $d(w) > 3$ or $w$ is in the queue, our algorithm proceeds as in the previous section.
Otherwise, there exist up to two degree chains $w, x_1,\ldots,x_p, y$ and $w,a_1,\ldots,a_q,b$ such that $x_1\neq v_k \neq a_1$.
Our algorithm iteratively tries to reduce all $D[x_i]$ and $D[a_i]$.
If it reaches $b$, $b$'s weight in the queue is set to $D[b]+h_t(b)$.
Analogously, if $y$ is reached, $y$'s weight is set to $D[y]+h_t(y)$.
If $y$ respectively $b$ are not reached, our algorithm does nothing.

@@

The problem with this straightforward approach is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
Thus, in the equivalent bidirectional Dijkstra search each direction would run on a different graph.
This breaks the usual bidirectional Dijkstra stopping criterion.
To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for either direction.
The forward search can skip nodes already settled by the backward search and vice versa.
However, this straightforward bidirectional A* still performs more work than a unidirectional A*~\cite{gh-cspas-05}.
Thus, on its own it is not a useful algorithm.
However, it can serve as basis for further algorithmic refinements.
The authors of~\cite{gh-cspas-05} refer to this approach as \emph{symmetric} bidirectional A*.

To obtain a bidirectional stopping criterion, the \emph{average potential} is proposed~\cite{gh-cspas-05}.
It combines a forward and a backward heuristic $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$ into a combined \emph{average heuristic} $h_{\overline{st}}$.
The idea is to use a common reduced graph, whose weights are the average weights of the individual reduced graphs.
Both searches run on the same common reduced graph.
This allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
Formally, $h_{\overline{st}}(v)$ is defined as $(\overrightarrow{h}_t(v) - \overleftarrow{h}_s(v))/2$.
The forward search uses $h_{\overline{st}}(v)$ as its heuristic.
The backward search uses $-h_{\overline{st}}(v)$.
Unfortunately, average potentials have two downsides.
First, evaluating the average potential requires evaluating both $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
Evaluating the average heuristic is therefore slower than evaluating just one of the composing heuristics.
Second, the bidirectional stopping criterion comes at the cost of worse estimates for each direction on its own.
$h_{\overline{st}}(v)$ is a worse estimate for $\dist(v,t)$ than $\overrightarrow{h}_t$.
Similarly, $-h_{\overline{st}}(v)$ is a worse estimate for $\dist(s,v)$ than $\overleftarrow{h}_s$\footnote{
To obtain an actual lower bound from this average heuristic, one has to add $\overleftarrow{h}_s(t)/2$ in the forward case and $\overrightarrow{h}_t(s)/2$ in the backward.
Adding any constant to a heuristic function does not change the reduced graph.
}.
The second downside can be partially mitigated through pruning with the composing heuristics.
When the forward search scans an edge $uv$ where $\overrightarrow{D}[u] + w(uv) + \overrightarrow{h}_s(v) > \mu$ holds, i.e. the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, it is not necessary to push $y$ into the queue.
The pruning rule for the backward search is analogous.

@@

To enable quick shortest path computations, we consider a two-phase setup with an additional off-line preprocessing phase before the on-line query phase.
The input to the preprocessing phase is a graph $G_\ell$ with lower bound weights $w_\ell$ and a node mapping function $\phi$.
$w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
We require that $w_q(u v)$ is greater or equal to the shortest distance $\dist_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
The output of the preprocessing is auxiliary data that enables an efficient heuristic function $h_t(x)$.
$h_t(x)$ is the exact distance from $\phi(x)$ to $\phi(t)$ in $G_\ell$.
In the applications considered in this paper, $w_\ell$ is always the freeflow travel time.

@@

To combine live traffic with predicted traffic, we define a modified travel time function $w_q$ that is then used as query weights.
Denote by $w_p(e,\tau)$ the predicted travel time along edge $e$ at moment $\tau$.
Further, $w_c(e)$ is the travel time according to current live traffic.
Finally, we denote by $\tau_{\mathrm{soon}}$ the moment when we switch to predicted traffic.
In our experiments, we set $\tau_{\mathrm{soon}}$ to one hour in the future.
We need to make sure that the modified travel time function fulfills the FIFO property.
For this reason, we cannot make a hard switch at $\tau_{\mathrm{soon}}$.
Our modified travel time function linearly approaches the predicted travel time. % with a bounded slope.
%
Formally, we set $w_q(e,\tau)$ to $w_c(e)$, if $\tau \leq \tau_{\mathrm{soon}}$.
Otherwise, we check whether $w_p(e,\tau_{\mathrm{soon}}) < w_c(e)$ is true.
If it is the case, we set $w_q(e,\tau)$ to $\max\{w_c(e)+(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
Otherwise, we set $w_q(e,\tau)$ to $\min\{w_c(e)-(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
% We then run the algorithm of Section~\ref{sec:predicted-traffic} with $w_q$ as query weight.
In our implementation, we do not modify the representation of $w_p$ but evaluate the formulas above at each travel time evaluation.
We set $w_\ell$ to the freeflow travel time.

@@

This setup provides an alternative way of integrating live traffic compared to the approach described in Section~\ref{sec:live-predicted-traffic}.
Changes to the live traffic situation are handled by changing $w_\ell$ via a customization.
Additional features such as turn restrictions can be implemented by adjusting $w_q$.
For severe traffic events, the three-phase approach will be significantly faster than the one described in Section~\ref{sec:live-predicted-traffic}.
However, it comes at the expense of a more complex algorithm.
Further, when integrating predicted traffic, CCH-Potentials will not have much advantage over CH-Potentials because the heuristic must be a lower bound for the entire time considered and not only the current situation.

@@

When looking at total running times for a single many-to-one query including selection times, Lazy RPHAST delivers very competitive performance and is as fast if not faster than all algorithms evaluated in~\cite{delling_et_al:OASIcs:2011:3266}.
For example, Delling et al. report an average RPHAST running time of 1.97\,ms for selection and query combined for $|B| = 2^{14}$ and 28.52\,ms for $|B| = 2^{20}$ on DIMACS Europe.
Lazy RPHAST (on a CH) takes 0.96\,ms and 10.76\,ms respectively, to compute the same distances.
The CCH variant is slightly slower (1.00\,ms and 14.41\,ms) but still significantly faster than RPHAST.
While the numbers are not perfectly comparable due to different benchmark machines, we can still conclude that Lazy RPHAST is a very useful extension to RPHAST.
It allows efficiently handling dynamic source sets and is competitive in the setting where both the target and the source set change between queries.
% This is of course not a fair comparison since the approaches in~\cite{delling_et_al:OASIcs:2011:3266} are geared towards having one expensive selection phase and than many faster queries.
% When computing distances from different sources towards the same target set, RPHAST will outperform Lazy RPHAST.
% However, whenever the target set may change between queries, Lazy RPHAST offers a simple and fast alternative.

@@

Table~\ref{tab:building_blocks} depicts the performance of unidirectional A* with different heuristics and optimizations.
We compare (C)CH-Potentials to three other heuristics.
The first heuristic is the \emph{Zero} heuristic where $h(v)$ is $0$ for all nodes $v$.
This corresponds to using Dijkstra's algorithm.
Secondly, we compare against our own implementation of ALT~\cite{gw-cppsp-05}.
We use 16 landmarks generated with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
Finally, we compare against a hypothetical \emph{Oracle-A*} heuristic.
This heuristic has instant access to a shortest distance array with respect to $w_\ell$, i.e. it is faster than the fastest heuristic possible in our model.
We fill this array before each query using a reverse Dijkstra search from the target node.
In this experiment, we do not account for the running time of Dijkstra's algorithm.
Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
(C)CH-Potentials compute the same distance estimates, but the heuristic evaluation has some overhead.
Comparing against Oracle-A* allows us to measure this overhead.
No other heuristic, which only has access to the preprocessing weights, can be faster than Oracle-A*.

@@

In this section, we investigate the performance of bidirectional A*.
We first evaluate different variants of bidirectional A* in Table~\ref{tab:bidir_pruning} and~\ref{tab:bidir_switching} then compare the best ones against unidirectional A* in Table~\ref{tab:bidir}.
Table~\ref{tab:bidir_pruning} studies the impact of our improved pruning and the low degree optimizations.
As observed in the previous section, enabling the low degree optimization achieves a speed-up by roughly a factor of three.
Symmetric bidirectional A* without our improved pruning has the worst performance.
Enabling the improved pruning significantly improves the performance of symmetric bidirectional A*.
For all heuristics except ALT, symmetric A* with the improved pruning has smaller search spaces than the average potential and very similar running times.
Without the low degree improvements, the improved symmetric variant is marginally faster and with the low degree improvements the average potential remains slightly faster.
This is due to the reduced impact of the heuristic evaluation overhead with the low degree optimizations.
Enabling the improved pruning for the average potential marginally reduces the search space size and slightly increases running times.

@@

In Table~\ref{tab:bidir} we investigate the effectiveness of bidirectional search compared to unidirectional search depending on the query weights.
Interestingly, only the zero heuristic and ALT consistently achieve speed-ups through bidirectional search on all query weights.
With $w_q = w_{\ell}$, unidirectional CH-Potentials are already optimal and only traverse the shortest path.
Here, bidirectional search will only introduce unnecessary overhead.
When query weights are scaled up uniformly, bidirectional search achieves some search space reduction, but it is not enough to significantly reduce running times due to the overhead of running a second search.
This changes drastically when the query weight increases are applied non-uniformly in the third scenario.
Here only weights with a speed less than 80\,kph were scaled up.
This touches only the beginning and end of most shortest paths between randomly chosen nodes.
The middle part of the shortest paths will typically use faster edges like highways.
In this case, bidirectional (C)CH-Potentials are a factor of two faster than the unidirectional variant.
The reason for this is that the search space of a unidirectional search expands greatly while exploring the end of the path to the target where the reduced weights are bad.
In contrast, the bidirectional searches meet in the middle of the shortest path where the reduced weights are close to zero and thus avoid this expansion.
This is also the reason why ALT behaves like this for all query weights.
By construction, the ALT heuristic has better reduced weights for edges which are on many shortest paths like highways.
Conversely, unimportant edges have bad reduced weights.
This makes bidirectional search so critical for ALTs performance.
In contrast, having a potential as tight as CH-Potentials makes bidirectional search in many scenarios unnecessary.
Bidirectional search for CH-Potentials only pays off when the reduced weights are bad around the terminals.

@@

Table~\ref{tab:applications} depicts the running times of (C)CH-Potentials in various applications, such as those described in Section~\ref{sec:extensions}.
We report speedups compared to extensions of Dijkstra's algorithm for each application respectively.
We start with the base case where $w_q = w_\ell$.
This is the problem variant solved by the basic CH algorithm.
CH achieves average query running times of 0.16\,ms on OSM Ger.
CH-Potentials are roughly four times slower but still achieve a huge speedup of 3795 over Dijkstra.
Such large speedups are typical for CH.
This shows that CH-Potentials gracefully converge toward CH in the $w_q = w_\ell$ special case.
On DIMACS Europe, the average degree is somewhat higher which makes the low-degree optimizations less effective which leads to more queue operations and a slightly slower running time.
